---
title: "Developing the R package for factor analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developing the R package for factor analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


Step 1: Introduction to Factor Analysis

Factor analysis is a statistical method used to identify underlying factors or dimensions that explain patterns in observed variables. It is commonly employed to reduce the dimensionality of data and uncover latent structures. The technique is based on the assumption that observed variables are influenced by a smaller number of unobservable factors.

There are two types of factor analysis, including confirmatory and exploratory factor analysis. In this project, we will focus on developing custom package including function for running both exploring both exploratory and confirmatory factor analysis. For, confirmatory factor analysis we will include include function for calculating factor loadings, cronbach alpha value and model fit. Whereas, for exploratory factor analysis, we will include function for calculating factor loadings for individuals items and cronbach alpha.



Assumptions of Factor Analysis:

Data Characteristics: Factor analysis assumes the absence of outliers, a sufficient sample size (where the number of cases exceeds the number of variables), and interval-level data measurement.

Statistical Assumptions: Perfect multicollinearity among variables should be absent. Additionally, although the model assumes linearity, non-linear variables can be transformed to meet this requirement.


Step 2: Background of the Data

The dataset used in this project involves the validation of the Hearing Handicap Inventory for Adults - Screening Version (HHIA-S) questionnaire in the Nepali language. This questionnaire is commonly used to assess hearing impairment among adult populations over short periods. While translations of the questionnaire exist in various languages, including Indian, Swedish, and Spanish, there is a lack of a Nepali version. Therefore, the data collection aimed to validate the translated questionnaire among both hearing-impaired and normal-hearing populations (N = 50(HI), N = 70(NH))

Both exploratory and confirmatory factor analyses were conducted on the dataset. Factor loadings, which represent the correlation between variables and factors, played a crucial role in these analyses.Factor loadings were calculated for each items in the dataset. A factor loading of 0.5 or higher typically indicates that the factor sufficiently captures the variance of the variable. 

Additionally, Cronbach's alpha value were calculated for each datasets. Cronbach's alpha measures the internal consistency of the scale or set of variables. A value closer to 1 indicates higher internal consistency

```{r}
#exploratory factor analysis
# Load required library, for exploratory factor analysis psych library is required (psych for both cronbach's alpha and EFA)
new_pkg <- c("tidyverse", "dplyr", "psych", "lavaan")
not_installed <- new_pkg[!new_pkg %in% rownames(installed.packages())]
if (length(not_installed) > 0) install.packages(not_installed) 
library(psych)
library(dplyr)
library(tidyverse)

# Load data for normal hearing individuals and hearing impaired population
normal_hearing_data <- read.csv("NH_DATA.csv")
hearing_impaired_data <- read.csv("HI_DATA.csv")

#Clear empty observations. After this process the observations of two data frames matches the numbers of participants (N = 50(HI), N = 70(NH))
normal_hearing_data <- normal_hearing_data %>%
  select(-X) %>%
  drop_na(Age)
hearing_impaired_data <- hearing_impaired_data %>%
  select(-X) %>%
  drop_na(Age)

# Define the questionnaire items for each group based on the column names
questionnaire_items <- c("E1", "E2", "S3", "E4", "S5", "S6", "E7", "S8", "E9", "S10")

# Define a function to calculate Cronbach's alpha and factor loadings for each group
calculate_alpha_and_loadings <- function(data, group_name) {
  # Select numeric variables
  numeric_data <- data[sapply(data, is.numeric)]
  
  # Perform factor analysis
  factor_analysis <- fa(numeric_data, nfactors = 1, rotate = "varimax")
  
  # Extract factor loadings
  factor_loadings <- factor_analysis$loadings
  
  # Print factor loadings
  print(paste("Factor Loadings for", group_name))
  print(factor_loadings)
  
  # Calculate Cronbach's alpha for all questionnaire items
  cronbach_alpha <- psych::alpha(numeric_data)$total$raw_alpha
  
  # Print Cronbach's alpha
  print(paste("Cronbach's Alpha for", group_name, ":", cronbach_alpha))
}

# Calculate Cronbach's alpha and factor loadings for normal hearing group
calculate_alpha_and_loadings(normal_hearing_data[questionnaire_items], "Normal Hearing")

# Calculate Cronbach's alpha and factor loadings for hearing impaired group
calculate_alpha_and_loadings(hearing_impaired_data[questionnaire_items], "Hearing Impaired")

```

```{r}
#Perform confirmatory factor analysis
#for CFA and Cronbach's alpha, both lavaan and psych are required (Cronbach's alpha  required psych)
# Load required libraries
library(lavaan)
library(psych)

# Load data for normal hearing individuals and hearing impaired population
normal_hearing_data <- read.csv("NH_DATA.csv")
hearing_impaired_data <- read.csv("HI_DATA.csv")

#Clear empty observations. After this process the observations of two data frames matches the numbers of participants (N = 50(HI), N = 70(NH))
normal_hearing_data <- normal_hearing_data %>%
  select(-X) %>%
  drop_na()
hearing_impaired_data <- hearing_impaired_data %>%
  select(-X) %>%
  drop_na()

# Define the questionnaire items for each group based on the column names
questionnaire_items <- c("E1", "E2", "S3", "E4", "S5", "S6", "E7", "S8", "E9", "S10")

# Define a function to perform confirmatory factor analysis (CFA) and calculate Cronbach's alpha and factor loadings for each group
perform_CFA <- function(data, group_name) {
  # Define the CFA model
  model <- '
    # Define latent variable
    factor =~ E1 + E2 + S3 + E4 + S5 + S6 + E7 + S8 + E9 + S10
  '
  
  # Perform CFA
  fit <- lavaan::cfa(model, data = data, std.lv = TRUE)
  print(fit)
  # Print standardized factor loadings
  print(paste("Standardized Factor Loadings for", group_name))
  print(inspect(fit, "std")$lambda)
  
  # Extract standardized factor loadings
  factor_loadings <- inspect(fit, "std")$lambda
  
  # Calculate Cronbach's alpha for all questionnaire items
  cronbach_alpha <- psych::alpha(data)$total$raw_alpha
  
  # Print Cronbach's alpha
  print(paste("Cronbach's Alpha for", group_name, ":", cronbach_alpha))
}

# Perform confirmatory factor analysis and Cronbach's alpha for normal hearing group
perform_CFA(normal_hearing_data[questionnaire_items], "Normal Hearing")

# Perform confirmatory factor analysis and Cronbach's alpha for hearing impaired group
perform_CFA(hearing_impaired_data[questionnaire_items], "Hearing Impaired")

```


Step 3: Develeloping  the Custom R Package
We create a custom R package that includes functions for both exploratory and confirmatory factor analysis. These functions will streamline the process of conducting factor analysis for users.



```{r setup}
library(FACTOR)
library(dplyr)
library(tidyverse)
```

Step 4: Preparing for analysis
We created a function `pick_check()` which intake two arguments: data and the type of test for checking whether the data is appropriate for FA. For example, we can input `"KMO"` to see the result of KMO test.
```{r}
pick_check("KMO", hearing_impaired_data[questionnaire_items])
```
```{r}
pick_check("Bartlett", normal_hearing_data[questionnaire_items])
```
The results states the data seem appropriate for FA.

Step 5: Performing Confirmatory Factor Analysis using the custom created package "FACTOR" (includes  factor loadings and Cronbach's alpha, and model fit)

Load dataset and clean empty observations
```{r}
normal_hearing_data <- read.csv("NH_DATA.csv")
hearing_impaired_data <- read.csv("HI_DATA.csv")
normal_hearing_data <- normal_hearing_data %>%
  select(-X) %>%
  drop_na()
hearing_impaired_data <- hearing_impaired_data %>%
  select(-X) %>%
  drop_na()
```


```{r, warning=FALSE}
questionnaire_items <- c("E1","E2","S3","E4","S5","S6","E7","S8", "E9", "S10")
```


Check CFA factor loadings

```{r}
model <- 
   " # Define latent variable
    factor =~ E1 + E2 + S3 + E4 + S5 + S6 + E7 + S8 + E9 + S10
    "
get_loadings_model(model, normal_hearing_data[questionnaire_items], "Normal Hearing")
get_loadings_model(model, hearing_impaired_data[questionnaire_items], "Hearing Impaired")
```
Check Cronbach's Alpha value

```{r,warning=FALSE}
model <- 
   " # Define latent variable
    factor =~ E1 + E2 + S3 + E4 + S5 + S6 + E7 + S8 + E9 + S10
    "
get_alpha_model(model, normal_hearing_data[questionnaire_items], "Normal Hearing")
get_alpha_model(model, hearing_impaired_data[questionnaire_items], "Hearing Impaired")
```

Check Model's fit

```{r}
model <- 
   " # Define latent variable
    factor =~ E1 + E2 + S3 + E4 + S5 + S6 + E7 + S8 + E9 + S10
    "
handy_fit(model, normal_hearing_data[questionnaire_items])
handy_fit(model, hearing_impaired_data[questionnaire_items])
```
Here are some fit statistics introduction and their reference levels (Maindal et al. 2012):

- `Chi-squared goodness-of-fit statistic` assesses the discrepancy between the sample and fitted covariance matrix (the null hypothesis is that the model fits the data. An insignificant test indicates good fit (`P-value` above 0.1). P-values of this model for two data sets do not indicate good fits. 

- `Comparative fit index (CFI)` assesses fit relative to a null model and ranges from 0 to 1 with values of 0.90–0.95 indicating acceptable and over 0.95 good fit. For this model, CFI indicates acceptable fits.

- `Tucker Lewis index (TLI)` adjusts for the number of model parameters and is interpreted as CFI.

- `Root mean square error of approximation (RMSEA)` expresses the lack of fit per degree of freedom of the model. Values are interpreted as follows: =<0.05 indicates very good, 0.05–0.08 good and >=0.10 poor fit. For this model, RMSEAs indicate poor fits.

- `Standardized root mean square residual (SRMR)` is the average of the differences between the observed and predicted correlations and has a range from 0 to 1. Values of <0.08 indicate good fit. For this model, SRMRs indicate good fits.

Step 7: Performing exploratory Factor Analysis using the custom created package "FACTOR" (includes  factor loadings and Cronbach's alpha, and model comparison)
```{r}
get_loadings(normal_hearing_data[questionnaire_items], 1, "Normal Hearing")
get_loadings(hearing_impaired_data[questionnaire_items], 1, "Hearing Impaired")

get_alpha(normal_hearing_data[questionnaire_items], 1, "Normal Hearing")
get_alpha(hearing_impaired_data[questionnaire_items], 1, "Hearing Imapired")
```

We might want to compare between model fits. By changing the second argument in in the functions in the chunk above could help with exploration. On the other hand, we could write two models and compare between the fits using `compare_fit()` function, which intake two models as the first two arguments and the data as the third argument.

For these datasets, we don't necessarily need to test other models due to the nature of the datasets. Here we just randomly write a second function as an example to demonstrate the function.
```{r}
model_test <- "
    # Define latent variable
factor1 =~ E1 + E2 + S3 + E4 + S5 + S6 + E7 + S8
factor2 =~ E9 + S10
  "
```

```{r}
compare_fit(model, model_test, hearing_impaired_data[questionnaire_items])
```

Step 6: Comparision of workflow using original Lavaan and psych package, and custom FACTOR package

The results of factor loadings, cronbach's alpha , and models fit for included datset is same using original package and custom made package. Hence, this proves we successfully created the custom package called "FACTOR". We bundled several functions from r packages focusing on FA and also included some created custom ones which helps us extract information/statistics in our FA.

References:

- Kline, P. (2014). An easy guide to factor analysis. Routledge.

- Flora, D. B., LaBrish, C., & Chalmers, R. P. (2012). Old and new ideas for data screening and assumption testing for exploratory and confirmatory factor analysis. Frontiers in Psychology, 3, 55. 

- Maindal, H. T., Sokolowski, I., & Vedsted, P. (2012). Adaptation, data quality and confirmatory factor analysis of the Danish version of the PACIC questionnaire. The European Journal of Public Health, 22(1), 31-36.

- Tavakol, M., & Dennick, R. (2011). Making sense of Cronbach's alpha. Int J Med Educ, 2, 53-55. https://doi.org/10.5116/ijme.4dfb.8dfd 

- Thompson, B. (2004). Exploratory and confirmatory factor analysis: Understanding concepts and applications. Washington, DC, 10694(000), 3.



